CrypTool Version 1.3 \\ Tutorial

### 2001-11-22 script-de/script-de.tex
no message

{\href{http://www.utm.edu/research/primes/proving.html\#intro}{\tt http://www.utm.edu/research/primes/proving.html\#intro}}.

@@

    ({\href{http://www.utm.edu/research/primes/programs/gallot/proths.html}
          {\tt http://www.utm.edu/research/primes/programs/gallot/proths.html}}).
  \item {\em Fermatzahlen} \index{Zahlen!Fermatzahlen}\index{Fermat!Fermatzahl}  $F(n) = 2^{2^n} + 1$:\\
    Wie oben erw"ahnt, schrieb Fermat diese Vermutung an Mersenne. Erstaunlicherweise w"are es f"ur
    ihn m"oglich gewesen, mit dem auf seinem kleinen Satz beruhenden negativen Primzahltest f"ur
    $n=5$  ein positives Ergebnis zu erhalten.
$$ \begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prim}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prim}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prim}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prim}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65.537 &  \mapsto {\rm ~prim}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 .294. 967. 297 = 641 \cdot
6.700.417 &  \mapsto {\rm ~NICHT~prim} !
   \end{array} $$
  \item Carmichaelzahlen: s.o.
  \item Pseudoprimzahlen: s.o.
  \item starke Pseudoprimzahlen: s.o.
  \item Idee aufgrund von \hyperlink{euclid}{Euklids Beweis} (unendlich viele Primzahlen)
        $p_1 \cdot p_2 \cdots p_n +1$:

@@

"Ahnlich wird die Anzahl der Primzahlen $PI(x)$ definiert, die den
Dachwert x nicht "ubersteigen:

@@

Man kann das auch so formulieren: Betrachtet man eine {\em zuf"allige} nat"urliche Zahl n, so sind die
Chancen, da"s diese Zahl prim ist, circa $1 / \ln(n)$. Nehmen wir zum Beispiel Zahlen in der Gegend von
$10^{16}$, so m"ussen wir ungef"ahr (durchschnittlich) $ 16 \cdot \ln 10 = 36,8 $
Zahlen betrachten, bis wir eine Primzahl finden.
Ein genaue Untersuchung zeigt: Zwischen $10^{16}-370$ und $10^{16}-1$ gibt es $10$
Primzahlen.

@@

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt http://www.math.Princeton.EDU/~arbooker/nthprime.html}}

@@

{\href{ftp://sable.ox.ac.uk/pub/math/cunningham/}{\tt ftp://sable.ox.ac.uk/pub/math/cunningham/}}

@@

\begin{itemize}
  \item Der deutsche Mathematiker \index{Goldbach Christian} Christian Goldbach (1690-1764) vermutete:
    Jede gerade nat"urliche Zahl gr"o"ser 2 l"a"st sich als die Summe zweier Primzahlen darstellen.
    Mit Computern ist die Goldbachsche Vermutung f"ur alle geraden Zahlen bis $10^8$ verifiziert, aber
    allgemein noch nicht bewiesen.
  \item Der deutsche \index{Riemann Bernhard} Mathematiker Bernhard Riemann (1826-1866) stellte eine bisher unbewiesene, aber auch nicht widerlegte (?)
Formel f"ur die Verteilung von
    Primzahlen auf, die die Absch"atzung weiter verbessern w"urde.
\end{itemize}

@@

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{http://www.math.Princeton.EDU/~arbooker/nthprime.html}}
\vskip +7 pt Ausgabe der n-ten Primzahl

@@

\begin{enumerate}
   \item \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }
   \item \href{http://www.utm.edu/research/primes/programs/gallot/proths.html}{\tt http://www.utm.edu/research/primes/programs/gallot/proths.html}
   \item \href{http://www.mersenne.org/prime.htm}{\tt http://www.mersenne.org/prime.htm }
   \item \href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html }
   \item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt http://www.eff.org/coop-awards/prime-release1.html }
%   \item \href{http://www.octocad.demon.co.uk/mersenne/prime.htm}{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm }
   \item \href{http://www.informatik.tu-darmstadt.de/TI/LiDIA/}{\tt http://www.informatik.tu-darmstadt.de/TI/LiDIA/ }
   \item \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt http://www.math.Princeton.EDU/~arbooker/nthprime.html }
   \item \href{http://www.utm.edu/research/primes/notes/by_year.html}{\tt http://www.utm.edu/research/primes/notes/by\_year.html }
   \item \href{ftp://sable.ox.ac.uk/pub/math/cunningham/}{\tt ftp://sable.ox.ac.uk/pub/math/cunningham/ }
\end{enumerate}

>>>>

{\href{http://www.utm.edu/research/primes/prove/index.html}{\tt
http://www.utm.edu/research/primes/prove/index.html}}.

@@

    ({\href{http://www.prothsearch.net/index.html}
          {\tt http://www.prothsearch.net/index.html}}).
  \item {\em Fermatzahlen} \index{Zahlen!Fermatzahlen}\index{Fermat!Fermatzahl}  $F(n) = 2^{2^n} + 1$:\\
    Wie oben erw"ahnt, schrieb Fermat diese Vermutung an Mersenne. Erstaunlicherweise w"are es f"ur
    ihn m"oglich gewesen, mit dem auf seinem kleinen Satz beruhenden negativen Primzahltest f"ur
    $n=5$  ein positives Ergebnis zu erhalten.
$$ \begin{array}{lll}
F(0) = 2^{2^0} + 1  = 2^1 + 1 & = 3 &   \mapsto {\rm ~prim}  \\
F(1) = 2^{2^1} + 1  = 2^2 + 1 & = 5 &   \mapsto {\rm ~prim}  \\
F(2) = 2^{2^2} + 1  = 2^4 + 1 & = 17 &  \mapsto {\rm ~prim}  \\
F(3) = 2^{2^3} + 1  = 2^8 + 1 & = 257 & \mapsto {\rm ~prim}  \\
F(4) = 2^{2^4} + 1  = 2^{16} + 1 &  = 65.537 &  \mapsto {\rm ~prim}  \\
F(5) = 2^{2^5} + 1  = 2^{32} + 1 &  = 4 .294. 967. 297 = 641 \cdot
6.700.417 &  \mapsto {\rm ~NICHT~prim} !
   \end{array} $$
  \item Carmichaelzahlen: s.o.
  \item Pseudoprimzahlen: s.o.
  \item starke Pseudoprimzahlen: s.o.
  \item Idee aufgrund von \hyperlink{euclid}{Euklids Beweis} (unendlich viele Primzahlen)
        $p_1 \cdot p_2 \cdots p_n +1$:

@@

"Ahnlich wird die Anzahl der Primzahlen $PI(x)$ definiert, die den
Dachwert $x$ nicht "ubersteigen:

@@

Man kann das auch so formulieren: Betrachtet man eine {\em zuf"allige} nat"urliche Zahl $n$, so sind die
Chancen, da"s diese Zahl prim ist, circa $1 / \ln(n)$. Nehmen wir zum Beispiel Zahlen in der Gegend von
$10^{16}$, so m"ussen wir ungef"ahr (durchschnittlich) $ 16 \cdot \ln 10 = 36,8 $
Zahlen betrachten, bis wir eine Primzahl finden.
Ein genaue Untersuchung zeigt: Zwischen $10^{16}-370$ und $10^{16}-1$ gibt es $10$
Primzahlen.

@@

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt
http://www.math.Princeton.EDU/\~arbooker/nthprime.html}}

@@

{\href{http://www.cerias.purdue.edu/homes/ssw/cun}{\tt
http://www.cerias.purdue.edu/homes/ssw/cun}}

@@

\begin{itemize}
  \item Der deutsche Mathematiker \index{Goldbach Christian} Christian Goldbach (1690-1764) vermutete:
    Jede gerade nat"urliche Zahl gr"o"ser $2$ l"a"st sich als die Summe zweier Primzahlen darstellen.
    Mit Computern ist die Goldbachsche Vermutung f"ur alle geraden Zahlen bis $10^8$ verifiziert, aber
    allgemein noch nicht bewiesen.
  \item Der deutsche \index{Riemann Bernhard} Mathematiker Bernhard Riemann (1826-1866) stellte eine bisher unbewiesene, aber auch nicht widerlegte (?)
Formel f"ur die Verteilung von
    Primzahlen auf, die die Absch"atzung weiter verbessern w"urde.
\end{itemize}

@@

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{http://www.math.Princeton.EDU/\~arbooker/nthprime.html}}
\vskip +7 pt Ausgabe der n-ten Primzahl

@@

\begin{enumerate}
   \item \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }
   \item \href{http://prothsearch.net/index.html}{\tt http://prothsearch.net/index.html}
   \item \href{http://www.mersenne.org/prime.htm}{\tt http://www.mersenne.org/prime.htm }
   \item \href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt http://reality.sgi.com/chongo/prime/prime\_press.html }
   \item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt http://www.eff.org/coop-awards/prime-release1.html }
%   \item \href{http://www.octocad.demon.co.uk/mersenne/prime.htm}{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm }
   \item \href{http://www.informatik.tu-darmstadt.de/TI/LiDIA/}{\tt http://www.informatik.tu-darmstadt.de/TI/LiDIA/ }
   \item \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt http://www.math.Princeton.EDU/\~arbooker/nthprime.html }
   \item \href{http://www.utm.edu/research/primes/notes/by_year.html}{\tt http://www.utm.edu/research/primes/notes/by\_year.html }
   \item \href{http://www.cerias.purdue.edu/homes/ssw/cun}{\tt http://www.cerias.purdue.edu/homes/ssw/cun }
\end{enumerate}

### 2001-11-22 script-gb/script-gb.tex
no message

\documentclass[a4paper,11pt]{article}
\usepackage[german]{babel}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{makeidx}
\makeindex
\hypersetup{colorlinks=true}

\setlength{\textwidth}{460 pt}
\hoffset-0.7in
\newtheorem{definition}{Definition}
\newcommand{\href}[2]{#2}
\newcommand{\hypertag}[2]{#1}
\newcommand{\hyperlink}[2]{#2}
\newcommand{\hypertarget}[2]{#2}
\newtheorem{theorem}{Theorem}
\newenvironment{Proof}[1]
  {\textbf{Proof #1} \\}
  {\hfill$\Box$ \\}

@@

\title{ CrypTool Version 1.2.02 \\ Tutorial }

\author{
(c) Deutsche Bank AG 1998-2000 \\
Frankfurt am Main 
}

@@

You will receive information about the principles of symmetrical and 
asymmetrical encryption. A large section of the tutorial is dedicated to the 
fascinating topic of prime numbers. You will also obtain an overview of the 
mathematical ideas behind modern cryptography.

@@

As at CrypTool, the quality of the tutorial is enhanced through your suggestions 
and ideas for improvement. We look forward to your feedback.

\vskip +10 pt \noindent
You will find the current version of CrypTool under 
\href{http://CrypTool.fzi.de}{\tt http://CrypTool.fzi.de} or \newline 
\href{http://www.cryptool.de}{\tt http://www.CrypTool.de}.

{
\vskip + 10 pt \noindent
The contact persons for this free tool are listed in the readme file for 
CrypTool.}
}

\newpage
\tableofcontents
\newpage
\normalsize
\parindent0cm
\parskip4pt

@@

A further aim was to enable users to understand the cryptographic procedures 
implemented in the Deutsche Bank. In this way, using CrypTool as a reliable 
reference implementation of the various encryption procedures, you can test the 
encryption implemented in other programs.\par \vskip + 3pt

@@

The advantage of asymmetric procedures is the easy \index{key management} key 
management. Let's look again at a network with $n$ subscribers. In order to 
ensure that each subscriber can establish an encrypted connection to each other 
subscriber, each subscriber must possess a pair of keys. We therefore need $2n$ 
keys or $n$ pairs of keys. Furthermore, no secure channel is needed before 
messages are transmitted, because all the information required in order to 
communicate confidentially can be transmitted openly. In this case, you simply 
have to pay attention to the accuracy (integrity and authenticity) 
\index{Authenticity} of the public key. Disadvantage: Pure asymmetric procedures 
take a lot longer to perform than symmetric ones.

@@

In other words: each natural number other than $1$ can be written as a product 
of prime numbers in precisely one way, if we ignore the order of the factors. 
The factors are therefore unique (the {\em expression as a product of factors} 
is unique)! For example, $$ 60 = 2 \cdot 2 \cdot 3 \cdot 5 = 2^2\cdot 3^1 \cdot 
5^1 $$
And this --- other than changing the order of the factors --- is the only way in 
which the number $60$ can be factorised. If you allow numbers other than primes 
as factors, there are several ways of factorising integers and the uniqueness is 
lost: $$ 60 = 1 \cdot 60 = 2 \cdot 30 = 4 \cdot 15 = 5 \cdot 12 =6 \cdot 10 = 2 
\cdot 3 \cdot 10 = 2 \cdot 5 \cdot 6 = 3 \cdot 4 \cdot 5 = ... $$

The following section is aimed more at those familiar with mathematical logic: 
The 1st fundamental theorem only appears to be obvious. We can construct 
numerous other sets of numbers (i.e. other than positive whole numbers greater 
than 1), for which numbers in the set cannot be expressed uniquely as a product 
of the prime numbers of the set: In the set $M = \{1, 5, 10, 15, 20, ...\}$ 
there is no equivalent to the fundamental theorem under multiplication. The 
first five prime numbers of this sequence are $5, 10, 15, 20, 30$ (note: $10$ is 
prime, because $5$ is not a factor of $10$ in this set --- the result is not an 
element of the given basic set $M$). Because the following applies in $M$: $$ 
100 = 5 \cdot 20 = 10 \cdot 10 $$ and $5, 10, 20$ are all prime numbers in this 
set, the expression as a product of prime factors is not unique here.

@@

{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

@@

{\href{http://www.utm.edu/research/primes/notes/faq/p_n.html}{\tt 
http://www.utm.edu/research/primes/notes/faq/p\_n.html}}.

@@

The definition is similar for the number of prime numbers $PI(x)$ that do not 
exceed the maximum value x:

@@

We can also express this as follows: Let us consider a {\em random} natural 
number n. Then the probability that this number is prime is around $1 / \ln(n)$. 
For example, let us take numbers in the region of $10^{16}$. Then we must 
consider $ 16 \cdot \ln 10 = 36,8 $ numbers (on average) until we find a prime. 
A precise investigation shows: There are $10$ prime numbers between $10^{16}-
370$ and $10^{16}-1$.

@@

{\href{http://www.utm.edu/research/primes/proving.html\#intro}{\tt 
http://www.utm.edu/research/primes/proving.html\#intro}}.

@@

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{http://www.math.Pr
inceton.EDU/~arbooker/nthprime.html}}
\vskip +7 pt Output of the nth prime number

@@

\newpage\begin{thebibliography}{99999}
\bibitem[Bartholome1996]{Bartholome1996}  \index{Bartholome 1996}     A. 
Bartholomé, J. Rung, H. Kern, \\     Zahlentheorie für Einsteiger, Vieweg 1995, 
2. Auflage 1996.

@@

\section*{URLs}

\begin{enumerate}
   \item \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }
   \item 
\href{http://www.utm.edu/research/primes/programs/gallot/proths.html}{\tt 
http://www.utm.edu/research/primes/programs/gallot/proths.html}
   \item \href{http://www.mersenne.org/prime.htm}{\tt 
http://www.mersenne.org/prime.htm }
   \item \href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt 
http://reality.sgi.com/chongo/prime/prime\_press.html }
   \item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt 
http://www.eff.org/coop-awards/prime-release1.html }
\item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt 
http://www.octocad.demon.co.uk/mersenne/prime.htm }
   \item \href{http://www.informatik.tu-darmstadt.de/TI/LiDIA/}{\tt 
http://www.informatik.tu-darmstadt.de/TI/LiDIA/ }
   \item \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt 
http://www.math.Princeton.EDU/~arbooker/nthprime.html }
   \item \href{http://www.utm.edu/research/primes/notes/by_year.html}{\tt 
http://www.utm.edu/research/primes/notes/by\_year.html }
   \item \href{ftp://sable.ox.ac.uk/pub/math/cunningham/}{\tt 
ftp://sable.ox.ac.uk/pub/math/cunningham/ }
\end{enumerate}

@@

Because the term {\em group} is used differently in everyday language than in 
mathematics, we will, for reasons of completeness, begin by introducing the 
essential statement of the formal definition of a group:
\begin{itemize}
   \item A group is a non-empty set $G$ and an operation $+$ . The set $G$ is 
closed under the operation $+$. Regardless of which two elements $a, b$ from $G$ 
are taken, performing the operation on them gives an element in $G$ (i.e. 
$a+b=c$, and $c$ lies in $G$).
   \item For all elements $a, b$ and $c$ in $G$: $(a+b)+c = a+(b+c)$.
   \item There exists an element $e$ in $G$ that behaves neutrally with respect 
to the operation $+$. That means that for all a in the set G: a+e = e+a = a.
   \item For each element $a$ in $G$ there exists a so-called inverse element $-
a$ ($-a$ also lies in $G$) such that: $a+(-a) = (-a)+a = e$.
\end{itemize}
If also $a+b = b+a$ for all $a, b$ in $G$ then we call the group an {\em 
Abelian} group. An operation denoted as $+$ indicates an {\em additive} group; 
if the operation is denoted as $\cdot$, we speak of a {\em multiplicative} 
group.

@@

An elliptic curve over ${\mathbb Z}_p$ is defined by an equation of the 
following form: $$ y^2 \; ({\rm mod} \; p) = x^3 + ax + b ({\rm mod} \; p) $$ 
(thus: equality in the field ${\mathbb Z}_p$), where $a, b$ are in ${\mathbb 
Z}_p$ and $4a^3 + 27b^2$ mod $p$ is not equal to zero. For fixed chosen numbers 
$a$ and $b$ in ${\mathbb Z}_p$, this equation has the pair of solutions $$ {\bf 
E} = \left\{(x,y) \left| \begin{array}{c} x {\rm ~and~} y {\rm ~are~in~} 
{\mathbb Z}_p {\rm ~and~} \\ y^2  \equiv x^3 + ax + b \; ({\rm mod~} p) {\rm 
~and~} \\ 4a^3 + 27b^2 \not\equiv 0 \; ({\rm mod} p)\end{array} \right.    
\right\}, $$ i.e. the set ${\bf E}$ consists of all pairs $x$ and $y$ that are a 
solution (in ${\mathbb Z}_p$) of the above equation. It must be noted that the 
numbers $a, b$ and $p$ determine which pairs $(x,y)$ lie in the set ${\bf E}$. 
This means that $a, b$ and $p$ specify this set. The elements $(x,y)$ in ${\bf 
E}$ are called points on the elliptic curve. In addition, ${\bf E}$ has one more 
element $O$ (the so-called point in infinity). The set ${\bf E}$ is usually 
called an elliptic curve. We can now define an operation (also denoted as $+$, 
although it is not the standard/usual addition of real numbers) on two elements 
in ${\bf E}$ such that the operation delivers an element that also lies in ${\bf 
E}$. The set ${\bf E}$ is therefore closed under the operation $+$. We can show 
that ${\bf E}$ is a group. The neutral element of the group ${\bf E}$ is the 
point in infinity $O$. Thus, for every two points $(x_1,y_1)$ and $(x_2,y_2)$ on 
the elliptic curve ${\bf E}$, there exists a point $(x_3,y_3)$ on ${\bf E}$ such 
that the operation $+$ complies with the following: $(x_1,y_1) + (x_2,y_2) = 
(x_3,y_3)$. Under certain circumstances, these points may also be equal to the 
point in infinity. Thus, when we speak of a point $P$ on an elliptic curve ${\bf 
E}$, we mean that $P = (x,y)$ and $(x,y)$ lies in the set ${\bf E}$. Any two 
points on an elliptic curve specified by $a, b$ and $p$ can therefore be added 
and the result is a point that also lies on the same elliptic curve. We must 
note that ${\bf E}$ can have the following meanings:
\begin{itemize}
   \item the set ${\bf E}$ of solutions pairs $(x,y)$ for an equation including 
the point $O$ \\[-20 pt]
   \item the group ${\bf E}$ (with the operation \glqq addition of $(x_1,y_1)$ 
and $(x_2,y_2)$\grqq) \\[-20 pt]
   \item the elliptic curve ${\bf E}$ (which is actually the same as the group 
${\bf E}$) \\[-20pt]
 \end{itemize}

@@

For cryptography, the important fact is that, for very large numbers, it appears 
to be extremely difficult to use a given point $Q$ on an elliptic curve to 
determine which two points have to be added to obtain $Q$. For large numbers $a, 
b$ and $p$ ($p$, for example, has a length of more than $160$ bits), the 
computer can easily add the point $P$  $m$ times after another, i.e. to 
determine the point $P + P + ... in an incredibly short space of time (in a few 
fractions of a second) + P = Q$ ($m$ summands $P$). Rather than $P + P + ... + P 
= Q$ ($m$ summands $P$) we also write $mP = Q$. If we have a point $P$ and a 
point $Q$, which both lie on the same elliptic curve, no procedure is known that 
enables us --- within an acceptable space of time --- to determine the number 
$m$ (assuming it actually exists) for which $mP = Q$. This is referred to as the 
\glqq elliptic curve discrete logarithm problem\grqq\ (abbreviated to ECDLP).

@@

\printindex

>>>>

\documentclass[a4paper,11pt]{article}
% Da man keine Unterst"utzung der deutschen Sprache in diesem Fall braucht!!
% \usepackage[german]{babel}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{makeidx}
\makeindex
\hypersetup{colorlinks=true}

\setlength{\textwidth}{460 pt}
\hoffset-0.7in
\newtheorem{definition}{Definition}
%\newcommand{\href}[2]{#2}
\newcommand{\hypertag}[2]{#1}
%\newcommand{\hyperlink}[2]{#2}
%\newcommand{\hypertarget}[2]{#2}
\newtheorem{theorem}{Theorem}
\newenvironment{Proof}[1]
  {\textbf{Proof #1} \\}
  {\hfill$\Box$ \\}

@@

\title{ CrypTool Version 1.3 \\ Tutorial }

\author{
(c) Deutsche Bank AG 1998-2001 \\
Frankfurt am Main
}

@@

You will receive information about the principles of symmetrical and
asymmetrical encryption. A large section of the tutorial is dedicated to the
fascinating topic of prime numbers. XXX You will also obtain an overview of the
mathematical ideas behind modern cryptography.XXX

@@

\enlargethispage{0.5cm}
As at CrypTool, the quality of the tutorial is enhanced through your suggestions
and ideas for improvement. We look forward to your feedback.

\vskip +7 pt \noindent
You will find the current version of CrypTool under \newline
  \href{http://www.CrypTool.de}{\tt http://www.CrypTool.de},
  \href{http://www.cryptool.com}{\tt http://www.CrypTool.com} or
  \href{http://www.cryptool.org}{\tt http://www.CrypTool.org}.
{
\vskip + 7 pt \noindent
The contact persons for this free tool are listed in the readme file for
CrypTool.}
}

\newpage
\tableofcontents
\addcontentsline{toc}{section}{Contents}
\newpage
\normalsize
\parindent0cm
\parskip4pt

@@

A further aim was to enable users to understand the cryptographic procedures
implemented in the Deutsche Bank. XXX In this way, using CrypTool as a reliable
reference implementation of the various encryption procedures, you can test the
encryption implemented in other programs. XXX \par \vskip + 3pt

@@

The advantage of asymmetric procedures is the easy \index{Key
management} key management. Let's look again at a network with $n$
subscribers. In order to ensure that each subscriber can establish
an encrypted connection to each other subscriber, each subscriber
must possess a pair of keys. We therefore need $2n$ keys or $n$
pairs of keys. Furthermore, no secure channel is needed before
messages are transmitted, because all the information required in
order to communicate confidentially can be transmitted openly. In
this case, you simply have to pay attention to the accuracy
(integrity and authenticity) \index{Authenticity} of the public
key. Disadvantage: Pure asymmetric procedures take a lot longer to
perform than symmetric ones.

@@

In other words: each natural number other than $1$ can be written as a product
of prime numbers in precisely one way, if we ignore the order of the factors.
The factors are therefore unique (the {\em expression as a product of factors}
is unique)! For example, $$ 60 = 2 \cdot 2 \cdot 3 \cdot 5 = 2^2\cdot 3^1 \cdot
5^1 $$
And this --- other than changing the order of the factors --- is the only way in
which the number $60$ can be factorised. If you allow numbers other than primes
as factors, there are several ways of factorising integers and the uniqueness \hypertarget{uniqueness}{} is
lost: $$ 60 = 1 \cdot 60 = 2 \cdot 30 = 4 \cdot 15 = 5 \cdot 12 =6 \cdot 10 = 2
\cdot 3 \cdot 10 = 2 \cdot 5 \cdot 6 = 3 \cdot 4 \cdot 5 = ... $$

The following section is aimed more at those familiar with mathematical logic:
The 1st fundamental theorem only appears to be obvious \label{remFundTheoOfArithm}. We can construct
numerous other sets of numbers (i.e. other than positive whole numbers greater
than 1), for which numbers in the set cannot be expressed uniquely as a product
of the prime numbers of the set: In the set $M = \{1, 5, 10, 15, 20, ...\}$
there is no equivalent to the fundamental theorem under multiplication. The
first five prime numbers of this sequence are $5, 10, 15, 20, 30$ (note: $10$ is
prime, because $5$ is not a factor of $10$ in this set --- the result is not an
element of the given basic set $M$). Because the following applies in $M$: $$
100 = 5 \cdot 20 = 10 \cdot 10 $$ and $5, 10, 20$ are all prime numbers in this
set, the expression as a product of prime factors is not unique here.

@@

%{\href{http://www.octocad.demon.co.uk/mersenne/prime.htm }{\tt http://www.octocad.demon.co.uk/mersenne/prime.htm}}.

@@

{\href{http://www.prothsearch.net/index.html}{\tt
http://www.prothsearch.net/index.html}}.

@@

The definition is similar for the number of prime numbers $PI(x)$
that do not exceed the maximum value $x$:

@@

We can also express this as follows: Let us consider a {\em random} natural
number $n$. Then the probability that this number is prime is around $1 / \ln(n)$.
For example, let us take numbers in the region of $10^{16}$. Then we must
consider $ 16 \cdot \ln 10 = 36,8 $ numbers (on average) until we find a prime.
A precise investigation shows: There are $10$ prime numbers between $10^{16}-
370$ and $10^{16}-1$.

@@

{\href{http://www.utm.edu/research/primes/howmany.shtml}{\tt
http://www.utm.edu/research/primes/howmany.shtml}}.

@@

{\href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{http://www.math.Pr
inceton.EDU/\~arbooker/nthprime.html}} \vskip +7 pt Output of the
nth prime number

@@

\newpage\begin{thebibliography}{99999}
\addcontentsline{toc}{subsection}{Bibliography}
\bibitem[Bartholome1996]{Bartholome1996}  \index{Bartholome 1996}     A.
Bartholomé, J. Rung, H. Kern, \\     Zahlentheorie für Einsteiger, Vieweg 1995,
2. Auflage 1996.

@@

\section*{URLs} \addcontentsline{toc}{subsection}{URLs}

\begin{enumerate}
   \item \href{http://www.utm.edu/}{\tt http://www.utm.edu/ }
   \item
\href{http://prothsearch.net/index.html}{\tt
http://prothsearch.net/index.html}
   \item \href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm }
   \item \href{http://reality.sgi.com/chongo/prime/prime_press.html}{\tt
http://reality.sgi.com/chongo/prime/prime\_press.html }
   \item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt
http://www.eff.org/coop-awards/prime-release1.html }
\item \href{http://www.eff.org/coop-awards/prime-release1.html}{\tt
%http://www.octocad.demon.co.uk/mersenne/prime.htm }
%   \item \href{http://www.informatik.tu-darmstadt.de/TI/LiDIA/}{\tt
http://www.informatik.tu-darmstadt.de/TI/LiDIA/ }
   \item \href{http://www.math.Princeton.EDU/~arbooker/nthprime.html}{\tt
http://www.math.Princeton.EDU/\~arbooker/nthprime.html }
   \item \href{http://www.utm.edu/research/primes/notes/by_year.html}{\tt
http://www.utm.edu/research/primes/notes/by\_year.html }
   \item \href{http://www.cerias.purdue.edu/homes/ssw/cun}{\tt
http://www.cerias.purdue.edu/homes/ssw/cun }
\end{enumerate}

@@

Because the term {\em group} is used differently in everyday language than in
mathematics, we will, for reasons of completeness, begin by introducing the
essential statement of the formal definition of a group:
\begin{itemize}
   \item A group is a non-empty set $G$ and an operation $+.$ The set $G$ is
closed under the operation $+.$ Regardless of which two elements $a, b$ from $G$
are taken, performing the operation on them gives an element in $G$ (i.e.
$a+b=c$, and $c$ lies in $G$).
   \item For all elements $a, b$ and $c$ in $G$: $(a+b)+c = a+(b+c)$.
   \item There exists an element $e$ in $G$ that behaves neutrally with respect
to the operation $+$. That means that for all a in the set G: $a+e = e+a = a.$
   \item For each element $a$ in $G$ there exists a so-called inverse element $-
a$ ($-a$ also lies in $G$) such that: $a+(-a) = (-a)+a = e$.
\end{itemize}
If also $a+b = b+a$ for all $a, b$ in $G$ then we call the group an {\em
Abelian} group. An operation denoted as $+$ indicates an {\em additive} group;
if the operation is denoted as $\cdot$, we speak of a {\em multiplicative}
group.

@@

An elliptic curve over ${\mathbb Z}_p$ is defined by an equation of the
following form: $$ y^2 \; ({\rm mod} \; p) = x^3 + ax + b ({\rm mod} \; p) $$
(thus: equality in the field ${\mathbb Z}_p$), where $a, b$ are in ${\mathbb
Z}_p$ and $4a^3 + 27b^2$ mod $p$ is not equal to zero. For fixed chosen numbers
$a$ and $b$ in ${\mathbb Z}_p$, this equation has the pair of solutions $$ {\bf
E} = \left\{(x,y) \left| \begin{array}{c} x {\rm ~and~} y {\rm ~are~in~}
{\mathbb Z}_p {\rm ~and~} \\ y^2  \equiv x^3 + ax + b \; ({\rm mod~} p) {\rm
~and~} \\ 4a^3 + 27b^2 \not\equiv 0 \; ({\rm mod} p)\end{array} \right.
\right\}, $$ i.e. the set ${\bf E}$ consists of all pairs $x$ and $y$ that are a
solution (in ${\mathbb Z}_p$) of the above equation. It must be noted that the
numbers $a, b$ and $p$ determine which pairs $(x,y)$ lie in the set ${\bf E}$.
This means that $a, b$ and $p$ specify this set. The elements $(x,y)$ in ${\bf
E}$ are called points on the elliptic curve. In addition, ${\bf E}$ has one more
element $O$ (the so-called point in infinity). The set ${\bf E}$ is usually
called an elliptic curve. We can now define an operation (also denoted as $+$,
although it is not the standard/usual addition of real numbers) on two elements
in ${\bf E}$ such that the operation delivers an element that also lies in ${\bf
E}$. The set ${\bf E}$ is therefore closed under the operation $+$. We can show
that ${\bf E}$ is a group. The neutral element of the group ${\bf E}$ is the
point in infinity $O$. Thus, for every two points $(x_1,y_1)$ and $(x_2,y_2)$ on
the elliptic curve ${\bf E}$, there exists a point $(x_3,y_3)$ on ${\bf E}$ such
that the operation $+$ complies with the following: $(x_1,y_1) + (x_2,y_2) =
(x_3,y_3)$. Under certain circumstances, these points may also be equal to the
point in infinity. Thus, when we speak of a point $P$ on an elliptic curve ${\bf
E}$, we mean that $P = (x,y)$ and $(x,y)$ lies in the set ${\bf E}$. Any two
points on an elliptic curve specified by $a, b$ and $p$ can therefore be added
and the result is a point that also lies on the same elliptic curve. We must
note that ${\bf E}$ can have the following meanings:
\begin{itemize}
   \item the set ${\bf E}$ of solutions pairs $(x,y)$ for an equation including the point $O$
   \item the group ${\bf E}$ (with the operation \glqq addition of $(x_1,y_1)$ and $(x_2,y_2)$\grqq)
   \item the elliptic curve ${\bf E}$ (which is actually the same as the group ${\bf E}$)
 \end{itemize}

@@

For cryptography, the important fact is that, for very large numbers, it appears
to be extremely difficult to use a given point $Q$ on an elliptic curve to
determine which two points have to be added to obtain $Q$. For large numbers $a,
b$ and $p$ ($p$, for example, has a length of more than $160$ bits), the
computer can easily add the point $P$  $m$ times after another, i.e. to
determine the point $P + P + \cdots + P = Q$ in an incredibly short space of time (in a few
fractions of a second) ($m$ summands $P$). Rather than $P + P + \cdots + P
= Q$ ($m$ summands $P$) we also write $mP = Q$. If we have a point $P$ and a
point $Q$, which both lie on the same elliptic curve, no procedure is known that
enables us --- within an acceptable space of time --- to determine the number
$m$ (assuming it actually exists) for which $mP = Q$. This is referred to as the
\glqq elliptic curve discrete logarithm problem\grqq\ (abbreviated to ECDLP).

@@

\addcontentsline{toc}{section}{Index} \printindex
